My wonderful paper

### 2024-12-03 index.qmd
working on the linear regression application

---
title: "My wonderful paper"
abstract: "Unexpected results in data analysis can prompt researchers to examine data quality and analysis steps. While some researchers can typically diagnose issues effectively with a few checks, others may struggle to identify appropriate checks to diagnose the problem. [an example of check] These checks are often informal and difficult to trace and discuss in publications, resulting in others questioning the trustworthiness of the analysis. To address this, we formalize the informal checks into an *analysis plan* that encompasses the analysis steps and (the unit tests): one test for whether the result meets expectations and multiple tests for checking the analysis. We then present a procedure to assess the quality of these unit tests based on their accuracy and redundancy on simulated versions of the original data. The accuracy is assessed using binary classification metrics, *i.e.*, precision and recall, while redundancy is calculated using mutual information. This procedure can be used to conduct a sensitivity analysis, compare different analysis plans, and to identify the optimal cutoff point for the unit tests."
cite-method: natbib
bibliography: references.bib
documentclass: jds
format:
  pdf:
   keep-tex: true
   toc: true
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

The rest of the paper is organized as follows: @sec-plan describes the concept of analysis plan in detail. @sec-examples provides examples of analysis plan [more details]. (need another section here or before examples?) @sec-conclusion concludes the paper.

@@

* the `testthat` package [@testthat] provides the infrastructure for testing R code. It allows users to write unit tests for functions and packages.
* The `assertthat` package [@assertthat] helps to write assertion statements that are supposed to be true at a certain point in the code for defensive programming.

# Analysis plan {#sec-plan}

@@

Accuracy refers to a set of tests' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives. A false positive can indicate (caution or skepticism on checking the data), whereas a false negative suggests the tests may lack sensitivity to unexpected outcomes. To model the relationship between the plan and outcome expectation (binary-binary), a logic regression model is used [@ruczinski_logic_2003]. Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables to solve regression problems [more introduction on logic regression]. Compared to other tree-based methods or machine learning methods for binary-binary prediction, logic regression produces Boolean combinations, or meta rules, that combines unit tests to solve the prediction problem. [need rewrite here]. The performance of the tests can be evaluated using precision and recall metrics derived from the confusion matrix of the logic regression prediction

  * precision: the proportion of correctly identified unexpected results (true positives) out of all the predicted unexpected results (true positives + false positives)
  * recall: the proportion of correctly identified unexpected results (true positives) out of all the actual unexpected results (true positives + false negatives)

@@

This concept extends naturally to multiple variables through total correlation [ref], $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:

@@

```{r}
#| label: fig-logic-reg
#| fig-cap: Logic regression model fitted to the three unit tests (test1, test2, test3) and the outcome expectation (unexpect) as the response variable. The model suggests using an OR rule to combine test1 and test3 to predict the outcome expectation.
fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
plot(fit)
```

@@

Three examples are presented to illustrate how the concept of analysis plan can be applied to data analysis. [toy example]. @sec-linear-reg illustrates how constructing  the result universe in a linear regression model of PM10 on mortality can help understand the impact of sample size, model specification, and variable correlation structure on data analysis. [example three]

@@

Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality). Analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model from the literature. This is the *outcome expectation*. There are multiple factors that can affect the outcome expectation of linear regression, which here is called *plan expectation*, for example, 1) sample size, 2) model specification, and 3) correlation structure between variables. Adequate sample size is required to achieve the desired power to detect the significance of PM10 on mortality. Temperature is often an important confounder to consider in such study (add reference). From some domain knowledge, an analyst may expect that the significance of PM10 coefficient can be attained by adding temperature to the model. Analysts may also expect certain correlation structure between PM10, temperature, and mortality, and the distribution of each variable.

 * add a paragraph to describe the simulation process

```{r sim-linear-res}
# Generate correlation matrices 
corr_grid <- expand.grid(seq(-0.01, -0.001, by = 0.001), 
                         seq(-0.6, -0.2, 0.05), seq(0.2, 0.6, 0.05))

# Function to compute correlation matrix for each combination 
gen_corr_mtx <- function(r1, r2, r3) {   
  cor_matrix <- matrix(c(1, r1, r2,                          
                         r1, 1, r3,                          
                         r2, r3, 1), nrow = 3, byrow = TRUE)          
  # Store the matrix in the list   
  if (all(eigen(cor_matrix)$values > 0)) return(cor_matrix) 
  }

# Plan for parallel processing
corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) 
  }) 
corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
sample_size <- c(50, 100, 500, 1000) 
model <- c("mortality ~ pm10 + temp", "mortality ~ pm10") |> map(as.formula)

generate_data <- function(n, mtx, seed = 123) {   
  mu <- c(0, 0, 0)   
  data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
  U <- pnorm(data, mean = 0, sd = 1)   
  set.seed(seed)      
  tibble(mortality = qpois(U[,1], 182), # assume distribution
         pm10 = qgamma(U[,2], shape = 6, rate = 0.2),           
         temp = qnorm(U[,3], mean = 55, sd = 16))   
  }

res <- tibble(corr_mtx = corr_mtx) |> 
  mutate(id = row_number()) |> 
  crossing(sample_size, model) |> 
  rowwise() |>   
  mutate(data = list(generate_data(n = sample_size, mtx = corr_mtx)),          
         fit = list(summary(lm(model, data))$coefficients))  
```

@@

* how to systematically simulate data is still unknown, sensitivity of the simulation to the results

* plotting is a critical way to check data and they can still be frame into a unit test. it is a open problem to how to encode the visualization into the unit tests. Maybe a procedure like confirm plot (this looks alright to you) and then press the button to continue

* currently no automated way to generate unit tests. It is interesting to see the automation of generating unit tests, although it requires the inputs from experts across a wide array of common scenarios.

* There are cost and benefit on setting expectation on different granularity. At the lowest level, one may have a plan for each data entry and every data handling steps. This requires more work from the analysts and may not be practical in practice. For more complex analyses, analysts may divide the analysis into sections and set expectations for each. They can then focus on the specific sections flagged by the tests and sub-divide the sections to set expectation and diagnose the analysis in a hierarchical manner.

* (only mention it) Software testing relies on "oracles" to provide the expected output necessary for verifying test results. For example, to test whether the program correctly calculates 1 + 1, one need to supply the correct answer, 2. However, establishing this "correct" output can sometimes be challenging, where obtaining a solution may be difficult without the program itself. This situation leads to the oracle problem [@barr2014oracle]. In data analysis, the similar oracle problem can happen, as the "truth" of an outcome, the expectation, depends on the underlying theory or interpretation. For example, in a linear regression model, the significance of a coefficient may be expected or unexpected based on the theory, making it challenging for researcher with a different theory to assess the results and the analysis.

>>>>

---
title: "My wonderful paper"
abstract: "Unexpected results in data analysis can prompt researchers to examine data quality and analysis steps. While some researchers can typically diagnose issues effectively with a few checks, others may struggle to identify appropriate checks to diagnose the problem. [an example of check] These checks are often informal and difficult to trace and discuss in publications, resulting in others questioning the trustworthiness of the analysis. To address this, we formalize the informal checks into an *analysis plan* that encompasses the analysis steps and (the unit tests): one test for whether the result meets expectations and multiple tests for checking the analysis. We then present a procedure to assess the quality of these unit tests based on their accuracy and redundancy on simulated versions of the original data. The accuracy is assessed using binary classification metrics, *i.e.*, precision and recall, while redundancy is calculated using mutual information. This procedure can be used to conduct a sensitivity analysis, compare different analysis plans, and to identify the optimal cutoff point for the unit tests."
cite-method: natbib
bibliography: references.bib
documentclass: jds
format:
  pdf:
   keep-tex: true
   toc: true
   notebook-links: false
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

@@

The rest of the paper is organized as follows: @sec-plan describes the concept of analysis plan in detail. @sec-examples provides examples of analysis plan \[more details\]. (need another section here or before examples?) @sec-conclusion concludes the paper.

@@

-   the `testthat` package [@testthat] provides the infrastructure for testing R code. It allows users to write unit tests for functions and packages.
-   The `assertthat` package [@assertthat] helps to write assertion statements that are supposed to be true at a certain point in the code for defensive programming.

# xxxxx (not analysis plan) {#sec-plan}

@@

Accuracy refers to a set of tests' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives. A false positive can indicate (caution or skepticism on checking the data), whereas a false negative suggests the tests may lack sensitivity to unexpected outcomes. To model the relationship between the plan and outcome expectation (binary-binary), a logic regression model is used [@ruczinski_logic_2003]. Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables to solve regression problems \[more introduction on logic regression\]. Compared to other tree-based methods or machine learning methods for binary-binary prediction, logic regression produces Boolean combinations, or meta rules, that combines unit tests to solve the prediction problem. \[need rewrite here\]. The performance of the tests can be evaluated using precision and recall metrics derived from the confusion matrix of the logic regression prediction

-   precision: the proportion of correctly identified unexpected results (true positives) out of all the predicted unexpected results (true positives + false positives)
-   recall: the proportion of correctly identified unexpected results (true positives) out of all the actual unexpected results (true positives + false negatives)

@@

This concept extends naturally to multiple variables through total correlation \[ref\], $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:

@@

```{r}
#| label: fig-logic-reg
#| fig.width: 6
#| fig.heigh: 2
#| fig-cap: Logic regression model fitted to the three unit tests (test1, test2, test3) and the outcome expectation (unexpect) as the response variable. The model suggests using an OR rule to combine test1 and test3 to predict the outcome expectation.
fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
plot_logicreg(fit)
```

@@

A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality. The example demonstrates how the test process can be used to select cutoff value in the unit tests and how the test can be used to diagnose the unexpected results.

@@

Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality). From the literature, analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model after considering the temperature confounding \[TODO add reference\]. This expectation can be framed into a unit test: the test fails, labelled as 1, if the p-value of the PM10 coefficient is greater than 0.05. And we have devised three unit tests to consider the 1) sample size, 2) strength of correlation between mortality and PM10, and 3) strength of correlation between mortality and temperature. \[sample size matters because ..., correlation between mortality and PM10 matters because...\]. However, it is not clear how to set the cutoff value for these tests to determine a failure. Here a list of tests are created in @tbl-tests.

| the test fails if ... | 
|--------------| 
| sample size less than or equal to 200 |
| sample size less than or equal to 400 |
| sample size less than or equal to 600 |
| sample size less than or equal to 800 |
|  mortality-PM10 correlation less than -0.03 |
|  mortality-PM10 correlation less than -0.04 |
|  mortality-PM10 correlation less than -0.05 |
|  mortality-PM10 correlation less than -0.06 |
| mortality-temperature correlation greater than -0.3 | 
| mortality-temperature correlation greater than -0.35 | 
| mortality-temperature correlation greater than -0.4 | 
| mortality-temperature correlation greater than -0.45 |

: A list of unit tests created for the linear regression model of mortality on PM10 and temperature. The tests are based on the sample size, correlation between mortality and PM10, and correlation between mortality and temperature. The cutoff values for each test are specified to determine a failure. {#tbl-tests}

To generate replicate of the data, we first simulate the correlation matrix for the three variables: mortality, PM10, and temperature with the following vector:

-   mortality and PM10: -0.03 to -0.1
-   mortality and temperature: -0.2 to -0.4
-   PM10 and temperature: 0.2 to 0.4

We then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size. The multivariate normal distribution is transformed using the normal CDF, followed by the inverse CDF of the assumed distributions of the three variables. Various distributions are fitted to the variables, including poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$. AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit. The following distributions and parameters, estimated from the original data, are used to simulate the three variables:

-   mortality, negative binomial distribution with size = 74 and mu = 183
-   PM10, beta distribution with shape1 = 4.21 and shape2 = 11.67, then multiply by 100
-   temperature, weibull distribution with shape = 3.8 and scale = 61

```{r}
#| label: fig-dist-fit
#| fig-cap: QQ-plot of the distribution fit for mortality, PM10, and temperature based on the fitted distribution from the original data. The fitted distribution is compared to the observed data to assess the distribution fit. 
#| fig.height: 4
#| fig.width: 10
aa2 <- fitdistrplus::fitdist(pm10$mortality, "nbinom")
p1 <- tibble(fitted= rnbinom(558, size = aa2$estimate[1], mu = aa2$estimate[2]),
       observed= pm10$mortality) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq() +
  stat_qq_line() + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("mortality: \nNB(size = 74, mu = 183)")

aa6 <- fitdistrplus::fitdist(pm10$pm10[pm10$pm10 < 90]/100, "beta")
p2 <- tibble(fitted = rbeta(557, shape1 = aa6$estimate[1], shape2 = aa6$estimate[2]) * 100,
       observed = pm10$pm10[pm10$pm10 < 90]) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq() +
  stat_qq_line() + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("pm10: \n100 * beta(shp1 = 4.21, shp2 = 11.67)")

aa4 <- fitdistrplus::fitdist(pm10$temp, "weibull")
p3 <- tibble(fitted= rweibull(558, aa4$estimate[1], aa4$estimate[2]),
       observed= pm10$temp) %>%
  ggplot(aes(sample = fitted)) +
  stat_qq() +
  stat_qq_line() + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  ggtitle("temperature: \nweibull(shape = 3.8, scale = 61)")

p1 | p2 | p3
```

```{r}
# cor(pm10$pm10, pm10$mortality)
# cor(pm10$temp, pm10$mortality)
# cor(pm10$temp, pm10$pm10)
# Generate correlation matrices 
corr_grid <- expand.grid(seq(0.01, 0.15, by = 0.02),                         
                         seq(0.25, 0.5, by = 0.05),                           
                         seq(0.25, 0.45, by = 0.05))

# Function to compute correlation matrix for each combination 
gen_corr_mtx <- function(r1, r2, r3) {   
  # correlation between mortality and pm10 are negative - r1
  # correlation between mortality and temp are negative - r2
  cor_matrix <- matrix(c(1, -r1, -r2,                          
                         -r1, 1, r3,                          
                         -r2, r3, 1), nrow = 3, byrow = TRUE)          
  # Store the matrix in the list   
  if (all(eigen(cor_matrix)$values > 0)) return(cor_matrix) 
  }

# Plan for parallel processing
corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) }) 
corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
sample_size <- c(400, 600, 800) 
model <- c("mortality ~ pm10 + temp") |> map(as.formula)

generate_data <- function(n, mtx, seed = 123, outlier = FALSE) {   
  mu <- c(0, 0, 0)   
  set.seed(seed)  
  data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
  U <- pnorm(data, mean = 0, sd = 1)   
  tibble(mortality = qnbinom(U[,1], size = 74, mu = 183),           
         pm10 = qbeta(U[,2], shape1 = 4.21, shape2 = 11.67) * 100,           
         temp = qweibull(U[,3], shape = 3.8, scale = 61))   
}

res <- tibble(corr_mtx = corr_mtx) |> 
  mutate(id = row_number()) |> 
  crossing(sample_size, model) |> 
  rowwise() |>   
  mutate(data = list(generate_data(n = sample_size, mtx = corr_mtx)),          
         fit = list(summary(lm(model, data))$coefficients))

pm10_df <- res |>   
  mutate(     
    p_value = fit[2,4],     
    coef = fit[2,1],     
    mor_pm10_correlation = corr_mtx[1, 2],     
    mor_temp_correlation = corr_mtx[1, 3],     
    temp_pm10_yz_correlation = corr_mtx[2, 3],     
    fml = deparse(model) |> as.factor(),     
    unexpect = ifelse(p_value < 0.05, 0, 1),
    #unexpect = ifelse(between(coef, 0, 0.15), 0, 1),
    `smpl_size <= 200` = ifelse(sample_size <= 200, 1, 0),
    `smpl_size <= 400` = ifelse(sample_size <= 400, 1, 0),
    `smpl_size <= 600` = ifelse(sample_size <= 600, 1, 0),
    `smpl_size <= 800` = ifelse(sample_size <= 800, 1, 0),
    `cor(m, PM10) < -0.03` = ifelse(mor_pm10_correlation < -0.03, 1, 0),
    `cor(m, PM10) < -0.04` = ifelse(mor_pm10_correlation < -0.04, 1, 0),
    `cor(m, PM10) < -0.05` = ifelse(mor_pm10_correlation < -0.05, 1, 0),
    `cor(m, PM10) < -0.06` = ifelse(mor_pm10_correlation < -0.06, 1, 0),
    `cor(m, tmp) > -0.3` = ifelse(mor_temp_correlation > -0.3, 1, 0),
    `cor(m, tmp) > -0.35` = ifelse(mor_temp_correlation > -0.35, 1, 0),
    `cor(m, tmp) > -0.4` = ifelse(mor_temp_correlation > -0.4, 1, 0),
    `cor(m, tmp) > -0.45` = ifelse(mor_temp_correlation > -0.45, 1, 0)
  ) |>    
  ungroup()

```

::: {#fig-linear-reg-tree}

```{r}
#| fig.width: 8
#| fig.heigh: 3
fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`cor(m, tmp) > -0.45`, 
                                seed = 4, nleaves = 3)
plot_logicreg(fit) 
```
<!-- The caption with some LaTeX like this $\lambda$. -->

Logic regression model fitted to the twelve unit tests and the outcome expectation (unexpect) as the response variable. The model suggests the relationship:
($\text{cor}(\text{m}, \text{tmp}) > -0.35$ AND $\text{smpl\_size} \le 600$) OR $\text{cor}(\text{m},  \text{PM10}) < - 0.05$
:::

```{r}
#| label: tbl-linear-reg
library(kableExtra)
df <- list(
  tibble(.fitted = pm10_df$`smpl_size <= 600`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.04`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.45`, 
         unexpect = pm10_df$unexpect),
  augment(fit)) |>
  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
            mutate(overlapping = 0, independence = 1) |> 
            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id")

tbl <- tibble(tests = c(1,2,3,4)) |> bind_cols(df |> dplyr::select(-id)) 
tbl |> 
  kable(digits = 3, escape = FALSE,
        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual unit tests (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation), and the combined test rule 4: (sample size AND mortality-temperature correlation) OR mortality-PM10 correlation.")

```

A total of `r ncol(pm10_df) - ncol(res) - 7` tests with four cutoff values each for the three types of tests are used to fit the logic regression and calculate the precision and recall score for predictability of an unexpected outcome (p value \< 0.05). @fig-linear-reg-tree shows the logic regression tree from the fitted model and @tbl-linear-reg prints the numerical summary of three selected single test and their combined test found by the logic regression. The logic regression model is capable of picking up the best cutoff value for each type of tests and yields the following three tests:

-   sample size less than or equal to 600
-   mortality-temperature correlation greater than -0.35
-   mortality-PM10 correlation less than -0.05

The combined test rule, (mortality-tempearture correlation AND sample size) OR mortality-PM10 correlation, has the highest precision, recall, and independence, suggesting that it is the most effective test for diagnosing unexpected results in the linear regression model given our assumption on the data. The tree structure suggests obtaining the mortality-PM10 correlation is as important as obtaining a combined relationship between mortality-temperature correlation and sample size in diagnosing unexpected results in the linear regression model. And based on our assumption of the variable distribution from the data simulation, we can obtain a precision of 0.908 and a recall of 0.794 for the unexpected outcome. These provides check points for the analysts to diagnose the unexpected results and to improve the model.

@@

-   how to systematically simulate data is still unknown, sensitivity of the simulation to the results

-   plotting is a critical way to check data and they can still be frame into a unit test. it is a open problem to how to encode the visualization into the unit tests. Maybe a procedure like confirm plot (this looks alright to you) and then press the button to continue

-   currently no automated way to generate unit tests. It is interesting to see the automation of generating unit tests, although it requires the inputs from experts across a wide array of common scenarios.

-   There are cost and benefit on setting expectation on different granularity. At the lowest level, one may have a plan for each data entry and every data handling steps. This requires more work from the analysts and may not be practical in practice. For more complex analyses, analysts may divide the analysis into sections and set expectations for each. They can then focus on the specific sections flagged by the tests and sub-divide the sections to set expectation and diagnose the analysis in a hierarchical manner.

-   (only mention it) Software testing relies on "oracles" to provide the expected output necessary for verifying test results. For example, to test whether the program correctly calculates 1 + 1, one need to supply the correct answer, 2. However, establishing this "correct" output can sometimes be challenging, where obtaining a solution may be difficult without the program itself. This situation leads to the oracle problem [@barr2014oracle]. In data analysis, the similar oracle problem can happen, as the "truth" of an outcome, the expectation, depends on the underlying theory or interpretation. For example, in a linear regression model, the significance of a coefficient may be expected or unexpected based on the theory, making it challenging for researcher with a different theory to assess the results and the analysis.
